Generating Random Logic Programs Using Constraint Programming

### 2020-01-04 paper/paper.tex
Variable symmetry elimination still needs some work

\begin{itemize}
\item Construct matrix M for variables that don't participate in the head. Have
  a variable variablesThatParticipateInTheHead that separates the set of
  variables.
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form. Not doing this on purpose. Leaving for further work.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

@@

% TODO: will need to be updated
\begin{example}
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\constants{} = \{ a
  \}$, $\variables{} = \{X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be
  represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{}}$. The node's
  $\variable{arity} \in [0, \maxArity{}]$ is defined by a \variable{table}
  constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{name} & \text{if } \variable{name} \in
      \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = 0\footnote{Zero
      here represents the first variable or constant, regardless of what it is.}.
  \]
\end{constraint}

\begin{example}
  Let $\maxArity{} = 2$, $\predicates{} = [\mathsf{P}, \dots]$, $\arities{}
  = [1, \dots]$, and $X \in \variables{}$. Then the node representing atom
  $\mathsf{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \mathsf{P},\\
    \variable{arguments} &= [X, 0],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

@@

\section{Eliminating Variable Symmetries}
% TODO: the matrix should include two more columns for bodies

\begin{definition} We define an array \variable{occurrences} of $|\constants{}|
  + |\variables{}|$ sets such that, for all $t \in \constants{} \cup
  \variables{}$, $\variable{occurrences}[t]$ is a subset of $\{ 0, \dots,
  \maxArity{} - 1\}$ that stores the indices of \variable{arguments} at which
  $t$ is located. It is defined by the \variable{setsIntsChanneling} constraint
  equivalent to the following condition: for $i = 0, \dots, \maxArity{} - 1$ and
  $t \in \constants{} \cup \variables{}$,
  \[
    \variable{arguments}[i] = t \quad \iff \quad i \in \variable{occurrences}[t].
  \]
\end{definition}

\begin{definition}
  We define a $|\variables{}| \times 2$ matrix $\mathbf{M}$ with domain $[0,
  \maxArity{}]$ such that for $v = 0, \dots, |\variables{}| - 1$,
  \begin{align*}
    M_{v,0} &= |\variable{occurrences}[v]|.\\
    M_{v,1} &= \begin{cases}
      \min \variable{occurrences}[v] & \text{if } \variable{occurrences}[v] \ne \emptyset\\
      0\footnotemark & \text{otherwise.}
    \end{cases}
  \end{align*}
\end{definition}
\footnotetext{This could be any fixed value in the domain.}

@@

@@

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: turn this into lexChainLessEq. But before: update the counting section.

@@

% TODO: does it work? explain why it works
Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v = 0}^a |\constants{}|^{a-v} \sum_{k=1}^{\min\{|\variables{}|,
    a\}} p_k(a),
\]
where $p_k(a)$ denotes the number of ways to partition integer $a$ into exactly
$k$ parts.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
%      h_1 + \dots + h_{|\predicates{}|} = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{C}{h_i} H_{a_i}^{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n + k - 1}{k}.
\]

This is without any kind of cycle detection and without probabilities.
% TODO: describe what parameter values were used to test the model.

>>>>

\begin{itemize}
\item A constraint for logical equivalence. An algorithm to reduce each tree to
  some kind of normal form. Not doing this on purpose. Leaving for further work.
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\item Could investigate how uniform the generated distribution of programs is.
  Distributions of individual parameters will often favour larger values
  because, e.g., there are more 5-tuples than 4-tuples.
\item Inference options to explore. Logspace vs normal space. Symbolic vs
  non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
  Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
  fsdd, fbdd.
\item Mention the random heuristic. Mention that restarting gives better
  randomness, but duplicates become possible. Restarting after each run is
  expensive. Periodic restarts could be an option.
\end{itemize}

@@

@@

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{} \cup \{ \Box
    \}}$, where $\Box$ denotes a disabled argument (the reason why this value is
  needed will become clear in \cref{sec:variable_symmetry}). The node's
  $\variable{arity} \in [0, \maxArity{}]$ is defined by a \variable{table}
  constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{name} & \text{if } \variable{name} \in
      \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$, if $i \ge \variable{arity}$, then
  $\variable{arguments}[i] = \Box$, else $\variable{arguments}[i] \ne \Box$.
\end{constraint}

\begin{example}
  Let $\maxArity{} = 2$, $\predicates{} = [\mathsf{P}, \dots]$, $\arities{}
  = [1, \dots]$, and $X \in \variables{}$. Then the node representing atom
  $\mathsf{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \mathsf{P},\\
    \variable{arguments} &= [X, \Box],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

@@

\section{Eliminating Variable Symmetries} \label{sec:variable_symmetry}

Given any clause, we can permute the variables in it without changing the
meaning of the clause or the entire program. Thus, we want to fix an order on
variables to eliminate unnecessary symmetries. Informally, we can say that
variable $X$ goes before variable $Y$ if its first occurrence in either the head
or the body of the clause is before the first occurrence of $Y$.

\begin{definition}
  Head:

  size: $|\constants{}| + |\variables{}|$

  domain: $\{ 0, \dots, \maxArity{} - 1\}$

  condition: for $i = 0, \dots, \maxArity{} - 1$ and $t \in \constants{} \cup \variables{}$,
  \[
    \variable{arguments}[i] = t \quad \iff \quad i \in \variable{occurrences}[t].
  \]

  Body:

  size: $|\constants{}| + |\variables{}| + 1$

  domain: $\maxNumNodes{} \times \maxArity{}$

  condition: for $i = 0, \dots, \maxNumNodes{} - 1$, $j = 0, \dots, \maxArity{}
  - 1$, $t \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    \variable{treeValues}[i].\variable{arguments}[j] = t \quad \iff \quad i
    \times \maxNumNodes{} + j \in \variable{occurrences}[t]
  \]

  We define an array \variable{occurrences} of  sets such that, for all $t \in
  \constants{} \cup \variables{}$, $\variable{occurrences}[t]$ is a subset of
  that stores the indices of \variable{arguments} at which $t$ is located. It is
  defined by the \variable{setsIntsChanneling} constraint equivalent to the
  following condition:
\end{definition}

\begin{definition}
  We define a $|\variables{}| \times 2$ matrix $\mathbf{M}$ with domain $[0,
  \maxNumNodes{} \times \maxArity{}]$ such that for $v = 0, \dots, |\variables{}| - 1$,
  \begin{align*}
    M_{v,0} &= |\variable{occurrences}[v]|.\\
    M_{v,1} &= \begin{cases}
      \min \variable{occurrences}[v] & \text{if } \variable{occurrences}[v] \ne \emptyset\\
      \maxNumNodes{} \times \maxArity{} & \text{otherwise.}
    \end{cases}
  \end{align*}
\end{definition}

@@

\begin{example}
  Let $\maxArity{} = 4$, $\mathsf{P} \in \predicates{}$, $\constants{} = \{ a
  \}$, $\variables{} = \{X, Y, Z \}$. Then $\mathsf{P}(Z, Y, Z)$ would be
  represented as:
  \begin{align*}
    \variable{predicate} &= \mathsf{P},\\
    \variable{arguments} &= [Z, Y, Z, 0],\\
    \variable{arity} &= 3,\\
    \variable{occurrences} &= [\emptyset, \emptyset, \{ 1 \}, \{ 0, 2 \}],\\
    \mathbf{M} &= \begin{bmatrix}
      0 & 0\\
      1 & 1\\
      2 & 0
    \end{bmatrix}.
  \end{align*}
\end{example}

@@

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: turn this into lexChainLessEq

@@

% TODO: explain how it works
Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v = 0}^a |\constants{}|^{a-v} \sum_{k=1}^{\min\{|\variables{}|,
    a\}} p_k(a),
\]
where $p_k(a)$ denotes the number of ways to partition integer $a$ into exactly
$k$ parts.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{C}{h_i} H_{a_i}^{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n + k - 1}{k}.
\]

This is without any kind of cycle detection and without probabilities.
% TODO: describe what parameter values were used to test the model (or... not).
